Interisland movements of woodland caribou in Newfoundland

### 2020-02-19 paper/manuscript.Rmd
restructured a lot of ms: - changed up intro to explain what we did, and then discussed some hypotheses post-hoc -

## Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
2016) to avoid predators [@Bergerud_1985] and access islands during calving
(Bergerud et al. 1990). Swimming for quadrupedal terrestrial mammals can be
thought of as a set of trade-offs. Costs include high energy expenditure and
risk of drowning. Specifically, even though caribou are adept swimmers, the
energetic expenditure associated with swimming for quadrupedal mammals is
significantly higher than walking or running (Fish 1993), whiledrowning
is also possible (Miller & Gunn 1985). Despite a large number of coastal and
island caribou herds, only a handful of examples have documented caribou
swimming in the ocean (Table 1).

Here, we describe the visual observation of an adult male caribou swimming
between islands in the Fogo Island archipleago off the coast of Newfoundland,
Canada. During routine fieldwork on May 30, 2017, we observed an unmarked adult
male caribou swim between Perry and Eastern Indian Islands (Figure 1), a swim of
at least 470 m which took ~9 minutes (~52 m per minute). This observation
prompted us to query a large database of GPS radio-collared caribou. We
identified  86 remotely sensed swimming events between XX islands from three
individual caribou over a three-year period.

<!--DO WE NEED THIS SECTION?--> 
Caribou were introduced to Fogo Island (n = 26 animals introduced between
1964–67) and Change Island (n = 5 animals introduced in 1964) from the main
Island of Newfoundland as part of a series of translocations and introductions
throughout the province (Bergerud & Mercer 1989). While caribou populations are
in decline throughout Canada (Festa-Bianchet et al. 2011), as well as in
Newfoundland (Bastille-Rousseau et al. 2013), the Fogo Island herd has not
declined to the same extent and currently consists of approximately 300 animals
(Newfoundland and Labrador Wildlife Division, unpublished data). Since 2016 we
have been studying the Fogo Island caribou herd (Bonar et al. 2017; Bonar et al.
2018; Peignier et al. 2019). To date, we have fitted XX adult female caribou
with GPS collars (1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
Canada), which were programmed to collect relocation fixes every 2 hours (see
Bonar et al. 2018 for details).

We posit two potential mechanisms explaining the trade-offs associated with
swimming in the North Atlantic Ocean: forage limitation and predator avoidance.
For caribou living on small islands in the ocean, the same notions for why
caribou swim in freshwater may also explain why caribou swim in the ocean
(Leblond et al. 2016). Specifically, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007) have been proposed as potential
explanations.

<!--I don't know if we need this and if yes, where it will go - thoughts?--> 
The Fogo Island archipelago is located in Newfoundland, Canada (49°N, 54°W), and
comprises approximately ten islands. While Fogo Island is the largest island
(237.71 km^2^), the archipelago includes at least three larger islands: Western
Indian (77.6 km2), Eastern Indian (38.7 km2) and Change (XX km2) Islands, as
well as numerous smaller islands, including Blundon’s (1.18 km2), North Long
(1.01 km2), South Long (0.48 km2), Kate’s (1.64 km2), and Brother’s (1.59 km2)
Islands (see Figure 1). All islands in the Fogo Island archipelago consist
largely of coniferous and mixed forests of balsam fir (Abies balsamea), black
spruce (Picea mariana), and white birch (Betula payrifera) as well as bogs,
lakes, lichen and rocky barrens.

## Forage limitation
Forage limitation and over-grazing is a major issue for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
islands. Reduced access to foraging opportunities can result in higher
conspecific competition for resources and lowered reproductive success for adult
females (Schaefer et al. 2016). Forage scarcity has been proposed as a potential
reason why caribou move between arctic islands on the sea-ice in winter (Miller
et al. 1977).  For very small islands, over-grazing by newly arrived caribou may
rapidly limit forage (Bergerud et al. 1990). Without sea ice, movement from the
mainland to an island as well as between islands requires caribou to assess the
trade-off associated with swimming.

The underlying mechanism driving forage limitation and the trade-off associated
with swimming could be density-dependent habitat selection governed by the Ideal
Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free Distribution
theory suggests that habitat selection is density-dependent and that variation
in density between habitat patches leads to a fitness equilibrium (Bradbury et
al. 2015). Fitness in a habitat patch depends on density where, ideally, the
available resources on a habitat patch can sustain a specific number of
individuals. In a hypothetical example, when density of a given habitat patch
has exceeded the fitness equilibrium, animals are predicted to relocate and
settle new habitat patches to reach equilibrium. Islands therefore represent
distinct habitat patches with clear costs and benefits associated with swimming.
For caribou in the Fogo Island archipelago, the role of forage limitation in
animal habitat selection patterns and movement remains unknown, but it is
possible that forage limitation and increased conspecific competition influences
swimming between islands.

## Predator avoidance
Predator avoidance could contribute to swimming behaviour of caribou (Miller
2002). Although caribou co-evolved with wolves (*Canis lupus*) throughout their
range, wolves are extirpated from Newfoundland, including Fogo Island. Coyotes
(*Canis latrans*) and black bears (*Ursus americanus*) are the primary predators
of caribou in Newfoundland (Bastille-Rousseau et al. 2016). While predation by
coyotes on adult female caribou does occur (ref?), it is relatively rare and the
effect of predation on swimming behaviour is more likely the result of
encounters which result in a flight response. For example, during calving,
females may swim to small islands to give birth (Bergerud & Page 1987) or swim
to islands in freshwater lakes with their calves to avoid predation.

Woodland caribou are also well known to disperse to remote locations,
including islands, during calving (Bergerud & Page 1987; Cumming & Beange 1987).
Moving to remote locations is a viable anti-predator strategy for caribou
because it reduces detection and encounter rates by predators. For instance, the
use of shoreline habitat in summer was deemed an effective anti-predator calving
strategy because caribou avoided predators by using islands and peninsulas or
swimming as a flight response (Bergerud et al. 1990). The use of water, and by
association occasionally swimming, is therefore part of the fine-scale
interactions between caribou and their predators and the use of islands appears
to be an effective anti-predator strategy. Coyotes and black bears are
responsible for the majority of mortalities for neonate caribou calves
(Bastille-Rousseau et al. 2016), although predation can still occur after the
calving period (Lewis & Mahoney 2014). However, on Fogo Island, coyotes are the
only predator of caribou.

## Summary and conclusion
Caribou were introduced to Fogo Island in the 1960s and for several decades the
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that caribou began
to occupy other islands in the archipelago. We surmise that, as density
increased over time, competition among conspecifics and density-dependent
habitat selection resulted in expansion of the population to nearby islands.
Following density-dependent habitat selection theory, caribou should swim to new
islands when the average fitness of individuals on the starting island exceeds
the density-fitness equilibrium (Morris 1987). Animals would therefore be
predicted to swim to new islands (i.e. settle new habitat patches) to reach a
fitness equilibrium. While this is an ultimate explanation to a series of
proximate observations, it is possible for density-dependent habitat selection
to operate at such a fine-scale (Webber & Vander Wal 2018). Evidence from the
Sable Island horse (Equus ferus caballus) system corroborates our suggestion
that swimming among islands could be driven by density-dependent habitat
selection (van Beest et al. 2014). For horses, selection for high quality
habitat was highest when population density was low, but individuals tended to
settle in lower quality habitats as population density increased, following
Ideal Free Distribution theory (van Beest et al. 2014). Taken together, caribou
may follow density-dependent habitat selection (Wittmer et al. 2007) and as
population density increased, forage was depleted and competition increased.

Our observations add to the evidence that caribou can, and occaissionaly do,
swim in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002;
Jeffery et al. 2007; Ricca et al. 2012) have noted the unique nature of caribou
swimming in the ocean, we suggest that this behaviour is likely more common than
previously thought for caribou living on oceanic islands. We posit forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for why caribou swimming in the ocean while predator avoidance is a
proximate explanation. Although our inference is limited to observations, our
observations suggest that islands represent discrete foraging patches for
terrestrial animals that could influence fitness via increased foraging
opportunities and safety from predators.

## Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by a National
Sciences and Engineering Research Council (NSERC) Vanier Canada Graduate
Scholarship to QMRW, NSERC Canda Graduate Scholarship to JGH, and a NSERC
Discovery Grant to EVW. We respectfully acknowledge the territory in which data
were collected and analyzed as the ancestral homelands of the Beothuk, and the
island of Newfoundland as the ancestral homelands of the Mi’kmaq and Beothuk.

>>>>

## Introduction
Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration (Leblond et al.
2016) to avoid predators [@Bergerud_1985] and access islands during calving
(Bergerud et al. 1990). Even for adept swimmers like caribou, the
energetic expenditure of swimming for quadrupedal mammals is
significantly higher than walking or running (Fish 1993), while drowning
is also possible (Miller & Gunn 1985). Despite a large number of coastal and
island caribou herds, only a handful of examples have documented caribou
swimming in the ocean (Table 1).

The Fogo Island archipelago, off the coast of Newfoundland, Canada (49°N, 54°W),
is home to approximately 300 woodland caribou. During routine fieldwork on May
30, 2017, we observed an unmarked adult male caribou swim between Perry and
Eastern Indian Islands (Figure 1), a swim of at least 470 m which took ~9
minutes (~52 m per minute). This observation prompted us to investigate the
subset of GPS radio-collared caribou in the population (n = 29 adult females).
Using these remotely sensed relocations, we determined that this observation was
not unique and that several individuals engaged in this behaviour (details below).
<!--Is it better to just state our results here, or give this sort of teaser? Initially I had "86 events by 3 individuals" here, but then we ignore that for a couple paragraphs of context before getting back into it... which seemed weird-->

Fogo Island is the largest island in the archipelago (237.71 km^2^), but there
are at least three other large islands: Western Indian (77.6 km2), Eastern
Indian (38.7 km2) and Change (XX km2) Islands, as well as numerous smaller
islands, including Blundon’s (1.18 km2), North Long (1.01 km2), South Long (0.48
km2), Kate’s (1.64 km2), and Brother’s (1.59 km2) Islands (see Figure 1).
Habitats are similar across the archipelago, consisting largely of coniferous
and mixed forests of balsam fir (*Abies balsamea*), black spruce (*Picea
mariana*), and white birch (*Betula papyrifera*) as well as bogs, lakes, lichen
and rocky barrens.

Caribou were introduced to Fogo Island (n = 26 animals between
1964–67) and Change Island (n = 5 animals in 1964) from the main
Island of Newfoundland as part of a series of translocations and introductions
throughout the province (Bergerud & Mercer 1989). While caribou populations are
in decline throughout Canada (Festa-Bianchet et al. 2011), as well as in
Newfoundland (Bastille-Rousseau et al. 2013), the Fogo Island herd has not
declined to the same extent and currently consists of approximately 300 animals
(Newfoundland and Labrador Wildlife Division, unpublished data). Since 2016 we
have been studying the Fogo Island caribou herd (Bonar et al. 2017; Bonar et al.
2018; Peignier et al. 2019). To date, we have fitted 29 adult female caribou
with GPS collars (1240g, GPS 4400M; Lotek Wireless Inc., Newmarket, Ontario,
Canada), which were programmed to collect relocation fixes every 2 hours (see
Bonar et al. 2018 for details). We used GPS locations from these individuals to investigate oceanic swimming events.

We identified swimming events as two consecutive GPS locations from an
individual on different islands. This region typically experiences pack ice
during the winter, so that caribou may be able to travel between islands by
walking over the ice. We restricted our GPS data to the ice-free period of the
year (April 1 - November 25) to avoid this possibility. Using these criteria, we identified three adult female caribou that collectively engaged in 86 swimming events (Fig 1??).

Caribou may swim between islands in the ocean for similar reasons that drive caribou swimming in freshwater (Leblond et al. 2016). We present two such
explanations for these swimming events, forage limitation (Miller 2002) and
predator avoidance (Jeffery et al. 2007), and discuss why these swimming events appear to be relatively commonplace for only a small subset of the population.

## Forage limitation
Forage limitation and over-grazing is a major concern for caribou populations
(Schaefer et al. 2016; Zamin et al. 2017) and could explain movement between
islands. Conspecific competition for resources driven by reduced foraging opportunities can decrease reproductive success of adult females
(Schaefer et al. 2016). Forage scarcity has been proposed as a potential reason
for caribou moving between arctic islands on the sea-ice in winter (Miller et
al. 1977).  For very small islands, over-grazing by newly arrived caribou may
rapidly deplete forage (Bergerud et al. 1990). Without sea ice, movement from the
mainland to an island, or between islands, requires caribou to assess the
trade-off associated with swimming.

This trade-off can be understood as density-dependent habitat selection governed
by the Ideal Free Distribution (Morris 1987; Bradbury et al. 2015). Ideal Free
Distribution theory predicts a fitness equilibrium: when the density in a given
habitat patch has exceeded the optimum for fitness within that patch, animals
should relocate and settle new habitat patches, so that fitness is equal across
all patches (Bradbury et al. 2015). Islands act as discrete habitat patches in
this case, but the costs associated with swimming create a trade-off when moving
between patches. For caribou in the Fogo Island archipelago, the role of forage
limitation in animal habitat selection patterns and movement remains unknown,
but swimming between islands may in part be explained by forage limitation and
conspecific competition.

## Predator avoidance
Predator avoidance could also contribute to swimming behaviour of caribou
(Miller 2002). During calving, female caribou are known to swim to small islands
in freshwater lakes give birth, or with their calves after birth, to avoid
predation (Bergerud & Page 1987). The use of shoreline habitat in summer is
considered an effective anti-predator calving strategy by allowing for swimming
to islands or peninsulas as a predation avoidance flight response (Bergerud et
al. 1990). The use of water, and by association occasionally swimming, is
therefore part of the fine-scale interactions between caribou and their
predators and the use of islands appears to be an effective anti-predator
strategy.

Historic predators of caribou such as wolves (*Canis lupus*) and black bears (*Ursus americanus*) are not present in the Fogo Island archipelago; coyotes (*Canis latrans*) are the top predator in this region. Predation by
coyotes on adult caribou is relatively rare (ref?), so the potential predator effect on caribou swimming behaviour is likely to be a flight response to non-lethal encounters. Calves are susceptible to coyotes, particularly young calves, but the energetic costs to oceanic swimming are likely to be far greater for small calves.

<!-- did the swimming events occur during the calving season? were these three breeders/non-breeders? did any of these females make these journeys with their calves?  -->

## Individual variation 
Three of 29 GPS-collared female caribou swam between islands in the Fogo
archipelago. These swimming events were relatively frequent, on average occuring
once every 24 days within our date range. However, we identified zero swimming
events for the other 26 collared females. It remains unclear why this behaviour
is seemingly common among 10% of our marked population, while 90% never engaged
in it. Our initial direct observation of swimming was an adult male, but as we
only have GPS data for females, we are unable to make conclusions on the
prevalence of this behaviour among male caribou. It is theoretically possible
that these other females moved rapidly back and forth between islands within the
2 hour fix rate of our GPS collars, which we would be unable to detect. We
consider this highly unlikely given the range of residency times for caribou on
any given island (min - max: XX - XX).
<!-- What is the typical residency time? Is this a true statement or am I just making this up? -->

The two hypotheses presented above are generally thought to apply uniformly
across populations, but individual caribou could evaluate the costs and benefits
of swimming to another island differently. Potentially those three females,
compared to the other 26, considered swimming less costly or perceived
competition or predation to be of greater concern. The individual attributes
that might influence how animals vary in their assessments of these trade-offs
are unknown, and potential predictors such as age or personality traits are
unknown for these caribou.

## Summary and conclusion
For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals (Bergerud & Mercer 1989; Newfoundland
and Labrador Wildlife Division, unpublished data). During the 1990s, population
density reached 300 animals and anecdotal evidence suggests that [only then???]
caribou began to occupy other islands in the archipelago. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium (Morris 1987). While this is an ultimate
explanation to a series of proximate observations, it is possible for
density-dependent habitat selection to operate at such a fine-scale (Webber &
Vander Wal 2018). Evidence from the Sable Island horse (Equus ferus caballus)
system corroborates this hypothesis (van Beest et al. 2014). For horses,
selection for high quality habitat was strongest at low population densities,
but individuals tended to settle in lower quality habitats as population density
increased, following Ideal Free Distribution theory (van Beest et al. 2014).
Predator avoidance can also contribute to caribou swimming behaviour, but in
this situation it seems a less likely explanation given predation is primarily a
concern for young calves, which are the least likely to engage in this costly
behaviour.
<!-- this last bit is sort of me bullshitting, not sure it's actually defensible?-->

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean (Table 1). While past observations (e.g. Miller 1995, 2002; Jeffery
et al. 2007; Ricca et al. 2012) have noted the exceptional nature of caribou
swimming in the ocean, we suggest that this behaviour is likely more common than
previously thought for caribou living on oceanic islands. We posit forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for why caribou swimming in the ocean, with predator avoidance as a
potential proximate explanation. Although our inference is limited to
observations, our observations suggest that islands represent discrete foraging
patches for terrestrial animals that could influence fitness via increased
foraging opportunities and safety from predators. We also suggest that
individuals vary in their evaluation of costs and benefits of movement between
these habitat patches.

## Acknowledgements 
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided by L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, K. Lewis, C.
Doucet, W. Barney, F. Norman, R. Otto, J. Neville, P. Saunders, T. Porter, P.
Tremblett, S. Gullage, T. Hodder, D. Jennings, J. McGinn, M. Frampton, P.
McClement, and S. Mahoney for data collection, management, and logistics. We
thank T. Bergerud and S. Mahoney for their vision in initiating much of the work
on caribou in Newfoundland. Funding for this study was provided by the National
Sciences and Engineering Research Council (QMRW, JGH, EVW). We respectfully
acknowledge the territory in which data were collected and analyzed as the
ancestral homelands of the Beothuk, and the island of Newfoundland as the
ancestral homelands of the Mi’kmaq and Beothuk.
<!-- Simplified the funding statement, do we need to name the specific scholarships/grants? Eric opinion??-->
